---
title: "Data manipulation with dplyr"
author: "Thiago Balbo Batista"
year: "2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Transforming data with dplyr


### Requesting libraries:

```{r libraries, echo = FALSE}
library('babynames')
library('dplyr')
library('ggplot2')
library('readr')
library('tidyverse')
```

### Importing database:

```{r import, echo=FALSE}
counties <- read_csv("acs2015_county_data.csv")
```

### Viewing the database:

```{r glimpse, echo=FALSE}

view_counties <- glimpse(counties)

```



### Selecting variables:

```{r select, echo=FALSE}

s_counties <- counties %>%  select(State, County, TotalPop, Poverty, 
                                   PublicWork, PrivateWork, SelfEmployed)

```


### Arranging dataframe:

```{r arrange, echo=FALSE}

# Sort this dataframe by PublicWork variable:

s_counties %>%  arrange(desc(PublicWork))

```




### Filtering for condition:

```{r filter, echo=FALSE}

# Filter counties with population above five thousand in California:

s_counties %>%  filter(State == 'California', TotalPop > 5000)

```

### Filtering and arranging:


```{r arrange/filter, echo=FALSE}

# Filter counties with population above five thousand in Texas
# Sort in descending order of percentage of people employed in private work

s_counties %>%  filter(State == 'Texas', TotalPop > 5000) %>% arrange(desc(PrivateWork))

```


### Mutate example:


```{r mutate, echo=FALSE}

# add a new variable of the absolute number of people employed in public work
# sort in descending order

s_counties %>% mutate(public_workers = TotalPop * (PublicWork/100)) %>% arrange(desc(public_workers))

```


### Mutate - porcentage of women in census tracts:


```{r mutate, echo=FALSE}

# select columns state, county, TotalPop, men and women
# calculate the percentage of women in the US counties

counties_selected <- counties %>% select(State, County, TotalPop, Men, Women) %>% mutate(percentage_women = (Women/TotalPop)*100)

```


### Combining select, arrange, filter, mutate:


```{r mutate, echo=FALSE}

# select the variables State, County, TotalpOP, Men, Women
# calculate the fraction of men in county's population
# arrange proportion_men in descending order
# filter counties with population above 10000

counties_selected <- counties %>% select(State, County, TotalPop, Men, Women) %>% mutate(proportion_men = (Men/TotalPop)*100) %>%  arrange(desc(proportion_men)) %>% filter(TotalPop > 10000) 

counties_selected

```


## Aggregating data

### Counting counties by state

```{r count, echo=FALSE}

counties_selected <- counties %>% select(State, County, TotalPop, Citizen ) 

# count the number of counties by State:
# arrange number of counties in descending order:

counties_selected %>% count(State) %>% arrange(desc(n))


```
### Counting citizens by state

```{r count, echo=FALSE}

# count the number of citizens by State:
# arrange number of counties in descending order:

counties_selected %>% count(State, wt = Citizen) %>% arrange(desc(n))

```


### Mutating and counting:

```{r count, echo=FALSE}

counties_selected <- counties %>% select(State, County, TotalPop, Walk)

# count the absolute number of people that go walking to work:

counties_selected  %>% mutate(population_walk = TotalPop * (Walk/100)) %>% arrange(desc(population_walk))

```

### Summarizing:

```{r summarize, echo=FALSE}

counties_selected <- counties %>%
  select(County, TotalPop, Income, Unemployment)

# County with the smallest population

counties_selected %>% summarise(min_population = min(TotalPop, na.rm = TRUE),   max_unemployment = max(Unemployment, na.rm = TRUE), average_income = mean(Income,na.rm = TRUE))


```


### Summarizing by state:

```{r summarize, echo=FALSE}

counties_selected <- counties %>% select(State, County, TotalPop)

# Total population by state using group by:

counties_selected %>% group_by(State) %>% summarize(total_population = sum(TotalPop) )

```


### Selecting a county from each state:

```{r top_n, echo=FALSE}

# The county in each state with highest percentage of people that go to work walkng:

counties_selected <- counties %>% select(State, County, TotalPop, Citizen, Walk)

counties_selected %>% group_by(State) %>% top_n(1, Walk)

```


## Select and transforming data

### Arguments in select function:

```{r select, echo=FALSE}

glimpse(counties)


counties %>% select(Professional:Production) %>% arrange(desc(Service))


```


```{r ends_with(), echo=FALSE}

# select columns related to work and filter counties where at least 50% of people is engaged in public work

counties %>% select(State, County, TotalPop, ends_with('work')) %>% filter(PublicWork >= 50)


```

### Rename column:

```{r rename, echo=FALSE}

counties %>% count(State) %>% rename(num_counties = n)


```

### renaming a column as part of a select:

```{r rename, echo=FALSE}

counties %>%   select(State, County, poverty_rate = Poverty)

```



### Using transmute:

```{r rename, echo=FALSE}

counties %>% transmute(State, County, poverty_rate = Poverty)

```


### Choosing among the four verbs:

```{r rename, echo=FALSE}

# Change the name of the unemployment column
counties %>%
  rename(unemployment_rate = Unemployment)

# Keep the state and county columns, and the columns containing poverty
counties %>%
  select(State, County, contains("poverty"))

# Calculate the fraction_women column without dropping the other columns
counties %>%
  mutate(fraction_women = Women / TotalPop)

# Keep only the state, county, and employment_rate columns
counties %>%
  transmute(State, County, employment_rate = Employed / TotalPop)

```

## Case Study: The babynames Dataset

### Filtering and arranging for one year:

```{r babynames , echo=FALSE}

babynames <- babynames

 # Filter for the year 1990
 # Sort the number column in descending order 

babynames %>% filter(year == 1990) %>% arrange(desc(n))

```

### Using top_n babynames:

```{r top_n babynames , echo=FALSE}

babynames %>%
  # Find the most common name in each year
  group_by(year) %>% top_n(1, n)
  

```
### Visualizing names with ggplot2:

```{r ggplot2 , echo=FALSE}

selected_names <- babynames %>%
  # Filter for the names Steven, Thomas, and Matthew 
  filter(name %in% c('Steven', 'Thomas', 'Matthew'))
# Plot the names using a different color for each name
ggplot(selected_names, aes(x = year, y = n, color = name)) +
  geom_line()
  

```

### Finding the year each name is most common:


```{r most common name , echo=FALSE}

# Calculate the fraction of people born each year with the same name
babynames %>%
  group_by(year) %>%
  mutate(year_total = sum(n)) %>%
  ungroup() %>%
  mutate(fraction = n/year_total) %>%

# Calculate the fraction of people born each year with the same name
  group_by(name) %>%
  top_n(1, fraction) 

```
### Adding the total and maximum for each name:

```{r maximum for each name , echo=FALSE}

babynames %>%
  # Add columns name_total and name_max for each name
   group_by(name) %>% mutate(name_total = sum(n),name_max = max(n))

```

### Using ratios to describe the frequency of a name


```{r frequency of a name , echo=FALSE}

babynames_fraction <- babynames %>%
                      group_by(year) %>%
                      mutate(year_total = sum(n)) %>%
                      ungroup() %>%
                      mutate(fraction = n / year_total)

babynames_fraction %>%
  # Arrange the data in order of name, then year 
  arrange(name, year) %>%
  # Group the data by name
  group_by(name) %>%
  # Add a ratio column that contains the ratio between each year 
  mutate(ratio = fraction / lag(fraction))

```
### Biggest jumps in a name:

```{r biggest jump , echo=FALSE}

babynames_fraction <- babynames %>%
                      group_by(year) %>%
                      mutate(year_total = sum(n)) %>%
                      ungroup() %>%
                      mutate(fraction = n / year_total)

babynames_fraction %>%
  # Arrange the data in order of name, then year 
  arrange(name, year) %>%
  # Group the data by name
  group_by(name) %>%
  # Add a ratio column that contains the ratio between each year 
  mutate(ratio = fraction / lag(fraction))

```

